|

### 2024-02-19 HocketEnforcersReport.Rmd
Updated

Creating the Surv object
```{r}
library(survival)
#Defining the survival variable
dat$Surv <- Surv(dat$age,dat$dead)
```
Some plots.
```{r}
ggplot(aes(x = age,col = EF), data = dat) + geom_density()
ggplot(aes(x = Wt,col = EF), data = dat) + geom_density()
ggplot(aes(x = Wt,y = age,col = EF), data = dat) + geom_point() + geom_smooth()
```

>>>>

table(dat$dead)
table(dat$EF)
table(dat$EP)

dat %>% group_by(EF) %>% summarize(n = n(), mnage = mean(age),seage = sd(age)/sqrt(n),mnwt = mean(Wt), sewt = sd(Wt)/sqrt(n))

mod <- lm(Wt ~ EF, data = dat)
predict(mod, newdata = data.frame(EF = c(FALSE,TRUE)), interval = "confidence")
#       fit      lwr      upr
# 1 196.2656 195.8592 196.6720
# 2 210.1511 208.4640 211.8381
dat$age <- as.numeric(dat$age)
mod <- lm(age ~ EF, data = dat)
predict(mod, newdata = data.frame(EF = c(FALSE,TRUE)), interval = "confidence")
#        fit      lwr      upr
# 1 47.60378 47.21229 47.99528
# 2 53.93270 52.30750 55.55789

t.test(dat$age~dat$EF)
t.test(dat$Wt~dat$EF)

dat %>% group_by(EP) %>% summarize(mnage = mean(age), n = n(), se = sd(age)/sqrt(n))
```

Creating the Surv object
```{r}
library(survival)
#Defining the survival variable
dat$Surv <- Surv(dat$age,dat$dead)
```
Some plots.
```{r}
ggplot(aes(x = age,col = EF), data = dat) + geom_density()
ggplot(aes(x = age,col = EP), data = dat) + geom_density()
ggplot(aes(x = Wt,col = EF), data = dat) + geom_density()
ggplot(aes(x = Wt,y = age,col = EF), data = dat) + geom_point() + geom_smooth()
```

### 2024-02-19 manuscript/paper.Rmd
Updated

# Methods {#sec:methods}
@Popkin2023 presents a matched cohort study with n = 6039 NHL players who players any games between the 1967-68 and the 2021-22 NHL seasons.  They then define a player as an enforcer in one of two ways:

1. Any player with at least 50 career fights (n = 331)
2. Any player who averaged at least 3 penalty minutes per game (n = 183)

Enforcers were then matched to non-enforcers on birth year, weight, height, [whagt else].  After matching, mortality rates between the two groups were computed.  Neither definition of enforcers yielded significant difference in the mortality rates between enforcers and non-enforcers.

@@

Greg doesn't think this conditioning on death is valid!  The point of matching is to balance the groups in regards to confounders.  However, after matching, you can't just look at a subset and compare betwee then two groups because you could end up with unbalanced sets!  And indeed we do see this!  For example, the weight in the dead enforcers group is on average higher than the weight of the dead non-enforcers group.  So we are comparing age at death of

The more I think about this, the less convinced I am that I'm "right" that this isn't a valid analysis.  Surely, they should have been doing survival analysis, but I can't quite articular why conditioning on the death and then looking at mean age is incorrect.  It just feels wrong to me and I'm sure there is a reason, but I am having trouble specifying exactly what the reason is.

@@

>>>>

# Data 
We obtained the two data sets used in @popkin2023: general table (n = 7432) and death table (n = 45).  After filtering for players who's first season was 1967 or greater, there are 6039 observations remaining (the exact number mentioned in @popkin2023).

@Popkin2023 presents a matched cohort study with n = 6039 NHL players who played any games between the 1967-68 and the 2021-22 NHL seasons.  (We removed 4 players with missing values for height leaving us with 6035 observations for analysis.)  They then define a player as an enforcer in one of two ways:

1. EF: Any player with at least 50 career fights (n = 331)
2. EP: Any player who averaged at least 3 penalty minutes per game (n = 183)

(Note: In our analysis I only have 182 players who are defined as EP)

While @popkin2023 only looks at a total of 45 deceased players  (21 enforcers and 24 controls), we scraped hockey-reference and identified 252 players who were included in the gen JAMA data who were deceased as of May 7, 2023.

## EDA
On average enforcers in this data set tended to be older than the non-enforcers (EF: 53.93 vs non-EF: 47.60 and EP: 52.50 vs non-EP: 47.81)

<!-- add a table here with summary statistics comparing the groups EF and EP.-->

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 Variable & EF = TRUE & EF = FALSE & p-value \\
 \hline
 Age (years) & `r summ$mnage[summ$EF]` (`r summ$seage[summ$EF]`) & `r summ$mnage[!summ$EF]` (`r summ$seage[!summ$EF]`) & `r ifelse(round(t.test(dat$age~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
 Weight (lbs) & `r summ$mnwt[summ$EF]` (`r summ$sewt[summ$EF]`) & `r summ$mnwt[!summ$EF]` (`r summ$sewt[!summ$EF]`) & `r ifelse(round(t.test(dat$Wt~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
  Height (in) & `r summ$mnht[summ$EF]` (`r summ$seht[summ$EF]`) & `r summ$mnht[!summ$EF]` (`r summ$seht[!summ$EF]`) & `r ifelse(round(t.test(dat$Ht~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$Ht~dat$EF)$p.value,2))`\\
 \hline
\end{tabular}
\end{center}

# Methods {#sec:methods}
@Popkin2023 presents a matched cohort study with n = 6039 NHL players who played any games between the 1967-68 and the 2021-22 NHL seasons.  They then define a player as an enforcer in one of two ways:

1. EF: Any player with at least 50 career fights (n = 331)
2. EP: Any player who averaged at least 3 penalty minutes per game (n = 183)

Enforcers were then matched to non-enforcers on birth year, weight, height, etc.  After matching, mortality rates between the two groups were computed.  Neither definition of enforcers yielded significant difference in the mortality rates between enforcers and non-enforcers.  (Greg comment: After matching you CAN compare mortality rates and that comparison would be valid AT THAT GIVEN POINT when they are compared.  But a better analysis would be to use survival analysis.  The problem with mortality rate is that eventually everyone in both groups will be dead if you wait long enough and that comparison is meaningless. Survival times are much better comparison. )

@@

Greg doesn't think this conditioning on death is valid!  The point of matching is to balance the groups in regards to confounders.  However, after matching, you can't just look at a subset and compare between then two groups because you could end up with unbalanced sets!  And indeed we do see this!  For example, the weight in the dead enforcers group is on average higher than the weight of the dead non-enforcers group.

@@

Add EDA results.

Add Cox models.

```{r echo = FALSE}
library(survival)
#Defining the survival variable
dat$Surv <- Surv(dat$age,dat$dead)
```

```{r echo = FALSE, fig.cap="Kaplan Meier plot for EF" }
library(ggsurvfit)
survfit2(Surv ~ EF, data = dat) %>% 
  ggsurvfit() +
  labs(
    x = "Year",
    y = "Overall survival probability"
  ) + 
  add_confidence_interval()
```

```{r echo = FALSE, fig.cap="Kaplan Meier plot for EP" }
library(ggsurvfit)
survfit2(Surv ~ EP, data = dat) %>% 
  ggsurvfit() +
  labs(
    x = "Year",
    y = "Overall survival probability"
  ) + 
  add_confidence_interval()
```

Talk about the conditioning on death stuff.

Age difference statement in Popkin is over stated.
